---
title: "Results"
author: "Toph Allen"
date: "September 8, 2015"
output: html_document
---


```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```


```{r load-packages, message = FALSE, warning = FALSE}
library(plyr)
library(dplyr)
library(magrittr)
library(lubridate)
library(ggplot2)
library(devtools)

load_all()
data(sos_raw)
data(sos_cleaned)
```

```{r survclass-calculations}
sos_surveillance_class <- sos_cleaned %>%
  select(sosid,
         humans,
         animals,
         plants,
         syndromic) %>%
  mutate(humans = factor(humans,
                         levels = c("yes", "no", "neia"),
                         labels = c("Yes", "No", "NEIA"),
                         ordered = TRUE),
        animals = factor(animals,
                         levels = c("yes", "no", "neia"),
                         labels = c("Yes", "No", "NEIA"),
                         ordered = TRUE),
        plants = factor(plants,
                         levels = c("yes", "no", "neia"),
                         labels = c("Yes", "No", "NEIA"),
                         ordered = TRUE),
        # I'm removing "blank" values from the Syndromic column, because
        # I don't think the difference is useful for this table.
        syndromic = revalue(syndromic, c("blank" = "neia")),
        syndromic = factor(syndromic,
                         levels = c("yes", "no", "neia"),
                         labels = c("Yes", "No", "NEIA"),
                         ordered = TRUE))
```

```{r org-calculations}
org_df <- clean_org_type(sos_raw$org_type, return_type = "data.frame")
org_factor <- clean_org_type(sos_raw$org_type, return_type = "factor")
sosid <- sos_cleaned$sosid
sos_org <- data.frame(sosid, org_df, org_factor)

```

The final database contains information on `r nrow(sos_raw)` biosurveillance systems. Of these, `r sum(sos_surveillance_class$humans == "Yes")` study humans, `r sum(sos_surveillance_class$animals == "Yes")` study animals, and `r sum(sos_surveillance_class$plants == "Yes")` study plants. We found `r sum(sos_surveillance_class$syndromic == "Yes")` systems which conduct syndromic surveillance and `r sum(sos_surveillance_class$syndromic == "No")` which did not, but were unable to find syndromic surveillance information for `r sum(sos_surveillance_class$syndromic == "NEIA")` systems. Government entities were involved in `r sum(sos_org$gov)` surveillance systems, non-profits in `r sum(sos_org$np)`, and for-profits in `r sum(sos_org$fp)`. More than one of these organization types were involved in `r sum(apply(sos_org[2:4], 1, function(x) sum(x) > 1))` systems.

```{r survclass-table}
knitr::kable(summary(sos_surveillance_class[2:5]),
             col.names = c("Humans", "Animals", "Plants", "Syndromic"),
             caption = "Table 1. Classes of biosurveillance conducted. Note: for \"Syndromic\" variable, \"Blank\" and \"Not Found\" entries were combined.")

```

```{r table-summary}
library(reshape2)
ggsurvclass <- sos_surveillance_class %>%
  melt(id.vars = "sosid") %>%
  mutate(variable = factor(variable, labels = c("Humans", "Animals", "Plants", "Syndromic Surveillance")),
         value = factor(value, levels = c("Yes", "No", "NEIA"), ordered = TRUE)) %>%
  ggplot()

ggsurvclass + geom_bar(aes(x = value, stat = "bin"), position = "dodge") + facet_grid(. ~ variable) + theme_bw() + labs(x = NULL, y = "Number of Systems", title = "Types of Biosurveillance Conducted")

```

```{r org-plot}
ggorg <- sos_org %>%
  select(sosid, org_factor) %>%
  ggplot()
  
ggorg + geom_bar(aes(x = org_factor, stat = "bin"), position = "dodge") + theme_bw() + labs(x = "Classes of Organization", y = "Number of Systems", title = "Types of entities conducting biosurviellance")

```

```{r time-series}
sos_dates <- select(sos_cleaned, date_created, date_terminated, current)
sos_dates$active_interval <- interval(sos_dates$date_created, sos_dates$date_terminated)


sum_active_systems_for_year <- function(year, intervals) {
  sum(year %within% intervals, na.rm = TRUE)
}

# Create a data frame for a time-series plot
time_series <- data.frame(year = parse_date_time(1900:2015, orders = "y"))

time_series$number_active <- sapply(time_series$year,
                                    sum_active_systems_for_year,
                                    intervals = sos_dates$active_interval)

# Create a data frame for the creation-termination histogram.
date_hist <- data.frame(sosid, select(sos_dates, date_created, date_terminated)) %>%
  melt(id.vars = "sosid") %>%
  mutate(year = year(value)) %>%
  filter(year != 2015, year >= 1950) %>%
  group_by(year, variable) %>%
  summarize(count = n()) %>%
  dcast(year ~ variable)

# We have to append rows for years not in the data frame, otherwise our over-under line will do weird things.
years_not_in_df <- seq(1950, 2014)[!seq(1950, 2014) %in% date_hist$year]
date_hist <- rbind(date_hist, data.frame(year = years_not_in_df, date_created = NA, date_terminated = NA))
date_hist[is.na(date_hist)] <- 0

max_creation <- date_hist %>%
  filter(date_created == max(date_created)) %>%
  select(year)
max_termination <- date_hist %>%
  filter(date_terminated == max(date_terminated)) %>%
  select(year)
most_systems_year <- time_series %>%
  filter(number_active == max(number_active)) %>%
  select(year) %>%
  mutate(year = year(year)) # Convert the "year" variable to a "year" from a full "date."

```

We examined data collected on surveillance system termination and creation. Of the `r nrow(sos_raw)` surveillance systems in the database, `r sum(!is.na(sos_dates$active_interval))` have information on both year of creation and, unless currently active, year of termination. Surveillance system creation peaked in `r max_creation[1, ]`, tailing off after `r max_creation[2, ]`. Termination of active systems increased significantly after 2000, peaking in `r max_termination`. Accounting for system creation and termination, the largest concurrent number of active surveillance systems was `r max(time_series$number_active)`, occurring in `r most_systems_year`.

```{r date_hist-plot}
ggplot(data = date_hist, aes(x = year)) +
  geom_bar(aes(y = date_created), stat = "identity", fill = "#00BFC4") +
  geom_bar(aes(y = -date_terminated), stat = "identity", fill = "#F8766D") +
  geom_line(aes(y = date_created - date_terminated), size = 0.5) +
  theme_bw() +
  labs(x = "Year", y = "Number of Surveillance Systems", title = "Number of Surveillance Systems Created and Terminated")

```

```{r time-series-plots}
ggplot(time_series, aes(x = year, y = number_active)) + geom_line() + theme_bw() + labs(x = "Year", y = "Count of Active Systems", title = "Number of Active Surveillance Systems over Time")

```


```{r lifespan-plot}
active_period <- as.period(sos_dates$active_interval, unit = "years")
active_duration <- as.duration(sos_dates$active_interval)

active_years <- year(active_period)

sos_dates %<>%
  mutate(years_active = year(as.period(active_interval, unit = "years")))

xlabs <- c(seq(0, 40, by = 10), ">50")

ggplot(sos_dates, aes(x = pmin(sos_dates$years_active, 50), fill = current)) +
  # layer(geom = "bar", stat = "bin", breaks = seq(0, 50)) +
  geom_bar(stat = "bin", breaks = 0:50) +
  scale_x_continuous(labels = c(seq(0, 40, by = 10), ">50")) +
  labs(x = "Years Active", y = "Count", title = "Lifespan of Biosurveillance Systems") +
  theme_bw()

ggplot(sos_dates, aes(x = pmin(sos_dates$years_active, 50), fill = current)) +
  geom_bar(stat = "bin",
           binwidth=.1,
           breaks = seq(0, 50, by = 5),
           position = position_dodge(width = 4),
           alpha = 0.9) +
  guides(fill = guide_legend(title = "Currently Active")) +
  scale_x_continuous(labels = c(seq(0, 40, by = 10), ">50")) +
  labs(x = "Years Active", y = "Count", title = "Lifespan of Biosurveillance Systems") +
  theme_bw()

```



```{r time-series-by-org-type}
sos_dates$org_type <- clean_org_type(sos_raw$org_type, return_type = "factor")
time_series_org <- expand.grid(year = time_series$year, org_type = levels(sos_dates$org_type))
time_series_org %<>%
  group_by(org_type, year) %>%
  mutate(number_active = sum_active_systems_for_year(year,
                                                     sos_dates[sos_dates$org_type == org_type,
                                                               "active_interval"])) %>%
  ungroup() %>%
  filter(year(year) >= 1990)


ggplot(time_series_org, aes(x = year, y = number_active, fill = org_type, color = org_type, order = desc(org_type))) + 
  geom_line() + 
  theme_bw() +
  labs(x = "Year", y = "Count of Active Systems", title = "Active Surveillance Systems by Organization Class")

```


```{r time-series-by-syndromic}
sos_dates$syndromic <- clean_yes_no_neia(sos_raw$syndromic)
sos_dates$syndromic <- factor(sos_dates$syndromic, levels = c("yes", "no"))
sos_dates$syndromic <- revalue(sos_dates$syndromic, replace = c("yes" = "Syndromic", "no" = "Traditional"))
time_series_org <- expand.grid(year = time_series$year, syndromic = levels(sos_dates$syndromic))
time_series_org %<>%
  group_by(syndromic, year) %>%
  mutate(number_active = sum_active_systems_for_year(year,
                                                     sos_dates[sos_dates$syndromic == syndromic,
                                                               "active_interval"])) %>%
  ungroup() %>%
  filter(year(year) >= 1980)


ggplot(time_series_org, aes(x = year, y = number_active, fill = syndromic, color = syndromic, order = desc(syndromic))) + 
  geom_bar(position = "dodge", stat = "identity") + 
  theme_bw() +
  labs(x = "Year", y = "Count of Systems", title = "Syndromic versus Traditional Surveillance Systems")

ggplot(time_series_org, aes(x = year, y = number_active, fill = syndromic, color = syndromic, order = desc(syndromic))) + 
  geom_bar(position = "fill", stat = "identity") + 
  theme_bw() +
  labs(x = "Year", y = "Proportion of Systems", title = "Syndromic versus Traditional Surveillance Systems")


```



```{r max-systems-in-country}
max_systems_in_country <- sos_countries %>%
  select(-sosid) %>%
  summarise_each(funs = funs(sum)) %>%
  max()

```

The United States was covered by the largest number of surveillane systems (`r max_systems_in_country`). However, the number of surveillance systems per country is not by itself a meaningful measure of surveillance effort, as the United States also covers a larger area and a larger population than many other countries. Plotting the number of surveillance systems per capita for each country shows a different aspect of surveillance system distribution.

```{r subset-countries}
countries_current <- sos_countries %>%
  filter(sos_dates$current) %>%
  select(-sosid)

```


```{r country-population}
library(curl)

# We must save the ISO3 names because we'll lose them with summarise_each.
iso3_names <- names(countries_current)

by_country <- countries_current %>%
  summarise_each(funs(n = sum(.))) %>%
  t() %>% # Transpose it so that countries are rows.
  as.data.frame() %>%
  mutate(iso3 = iso3_names) %>%
  arrange(desc(V1))
names(by_country)[1] <- "n_systems"

pop_csv <- curl_download(url = "https://raw.githubusercontent.com/datasets/population/master/data/population.csv",
                         destfile = tempfile())
population <- read.csv(pop_csv, stringsAsFactors = FALSE)
population %<>%
  filter(Year == 2014) %>%
  select(iso3 = Country.Code, population = Value) 
by_country %<>%
  left_join(population) %>%
  mutate(systems_per_capita = n_systems / population)
```

```{r maps}
library(rworldmap)
library(RColorBrewer)

sos_map <- joinCountryData2Map(dF = by_country, joinCode = "ISO3", nameJoinColumn = "iso3", verbose = FALSE)

sos_map$log_n_systems <- log(sos_map$n_systems)
sos_map$log_n_systems
spplot(sos_map, zcol = "log_n_systems", main = "Count of Surveillance Systems")

sos_map$log_systems_per_capita <- log(sos_map$systems_per_capita)
spplot(sos_map, zcol = "log_systems_per_capita", main = "Surveillance Systems Per Capita")

sos_map$systems_per_100000 <- sos_map$systems_per_capita * 100000


sos_map$systems_pretty <- cut(sos_map$systems_per_100000,
                              breaks = exp(pretty(log(sos_map$systems_per_100000))), dig.lab = 2)

spplot(sos_map, zcol = "systems_pretty", col.regions = rev(brewer.pal(6, "YlGnBu")), main = "Surveillance Systems Per 10,000 People")

```


## One Health analysis

```{r one-health}
onehealth <- sos_surveillance_class[, c("humans", "animals", "plants")] %>%
  mutate_each(funs(revalue(., replace = c("Yes" = "TRUE", "No" = "FALSE")))) %>%
  mutate_each(funs(as.logical))

# Version treating NAs as FALSE
onehealth_nona <- onehealth
onehealth_nona[is.na(onehealth_nona)] <- FALSE

onehealth %<>%
  mutate(humans_only = humans & !(plants | animals),
         animals_only = animals & !(humans | plants),
         plants_only = plants & !(humans | animals)) %>%
  mutate(ha = humans & animals,
         pa = animals & plants,
         hap = humans & animals & plants)


onehealth_nona %<>%
  mutate(humans_only = humans & !(plants | animals),
         animals_only = animals & !(humans | plants),
         plants_only = plants & !(humans | animals)) %>%
  mutate(ha = humans & animals,
         pa = animals & plants,
         hap = humans & animals & plants)


knitr::kable(summary(onehealth))

knitr::kable(summary(onehealth_nona))

```

```{r other-crap}
sos_map@data %>%
  select(ISO3, systems_per_100000) %>%
  arrange(-systems_per_100000) %>%
  knitr::kable()



```